Probabilistic inference in dynamical systems

### 2024-04-06 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): add manual derivation of analytical solution Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

in \eqref{eq-dustardtstar} and \eqref{eq-dsstardtstar}, which contains the $6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$. It is trivial to solve this system analytically, but in more general cases this will not be possible. We can now proceed to simulate the system to characterize the possible solution sets and then perform probabilistic inference on the latent variables.

>>>>

in \eqref{eq-dustardtstar} and \eqref{eq-dsstardtstar}, which contains the 
$6 - 2 = 4$ variables and parameters: $u^{\ast}, s^{\ast}, t^{\ast}, \gamma^{\ast}$.

We'll solve the dimensionless system analytically, focusing on what can be 
understood about the system's dynamics.
Given generic initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, the 
first equation

$$
\frac{du^{\ast}}{dt^{\ast}} = 1 - u^{\ast},
$$

can be solved using separation of variables

$$
u^{\ast}(t^{\ast}) = C_1 e^{-t^{\ast}} + 1.
$$

$C_1$ is the constant determined by initial conditions, which can be shown to be $C_1 = u^{\ast}_0 - 1$, giving us

$$
u^{\ast}(t^{\ast}) = (u^{\ast}_0 - 1) e^{-t^{\ast}} + 1.
$$

This equation describes how the concentration of pre-mRNA, in units of the 
characteristic concentration scale given by $\alpha/\beta$, evolves over the 
timescale characterized by the splicing rate. The term $e^{-t^{\ast}}$ 
indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
exponentially over time.

Now, we substitute the expression for $u^{\ast}(t^{\ast})$ into the 
differential equation for mRNA concentration

$$
\frac{ds^{\ast}}{dt^{\ast}} = u^{\ast} - \gamma^{\ast} s^{\ast}
$$

which gives

$$
\frac{ds^{\ast}}{dt^{\ast}} = [(u^{\ast}_0 - 1) e^{-t^{\ast}} + 1] - \gamma^{\ast} s^{\ast}
$$

This equation is a non-homogeneous first-order linear ODE for $s^{\ast}$. 
To solve it, we can use an integrating factor $e^{\gamma^{\ast} t^{\ast}}$, 
which gives

$$
e^{\gamma^{\ast} t^{\ast}} \frac{ds^{\ast}}{dt^{\ast}} + \gamma^{\ast} e^{\gamma^{\ast} t^{\ast}} s^{\ast} = (u^{\ast}_0 - 1) e^{(\gamma^{\ast}-1) t^{\ast}} + e^{\gamma^{\ast} t^{\ast}}
$$

Integrating both sides with respect to $t^{\ast}$ yields $s^{\ast}(t^{\ast})$

$$
s^{\ast}(t^{\ast}) = C_2 e^{-\gamma^{\ast} t^{\ast}} + \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} e^{-t^{\ast}} + \frac{1}{\gamma^{\ast}}
$$

where $C_2$ is another integration constant determined by initial conditions.
Incorporating the initial condition  $s^{\ast}(0) = s^{\ast}_0$ into the solution for $s^{\ast}(t^{\ast})$ to determine $C_2$

$$
C_2 = s^{\ast}_0 - \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1} - \frac{1}{\gamma^{\ast}}.
$$

With $C_2$ determined, we define

$$
\xi = \frac{u^{\ast}_0 - 1}{\gamma^{\ast} - 1},
$$

so that the complete solutions for the dimensionless system are given by

\begin{align}
u^{\ast}(t^{\ast}) & = \underbrace{1}_{\text{Steady State}} + \underbrace{(u^{\ast}_0 - 1) e^{-t^{\ast}}}_{\text{Transient Component}},  \label{eq-ustar-dimless} \\
s^{\ast}(t^{\ast}) & = \underbrace{\frac{1}{\gamma^{\ast}}}_{\text{Steady State}} + \underbrace{\left( s^{\ast}_0 - \xi - \frac{1}{\gamma^{\ast}} \right) e^{-\gamma^{\ast} t^{\ast}} + \xi e^{-t^{\ast}}}_{\text{Transient Components}}.  \label{eq-sstar-dimless}
\end{align}

The **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$ \eqref{eq-ustar-dimless} quickly 
settles to its steady state value of 1, regardless of the initial concentration, 
reflecting the balance between transcription and splicing rates. Note again that this
does not refer to an arbitrary value of $1$ but rather represents the equivalence
class of ratios $\alpha/\beta$.

The **mRNA dynamics** $s^{\ast}(t^{\ast})$ \eqref{eq-sstar-dimless} reveals a 
slightly more complicated balance of forces. The first transient term shows an exponential 
decay influenced by $\gamma^{\ast}$, the degradation rate relative to the splicing rate. 
The second transient term reflects how the 
initial deviation in pre-mRNA concentration from its steady state affects mRNA 
levels over time. The steady state of $s^{\ast}$ simply reflects the balance
of the splicing rate, which is the only source of mRNA production in this model,
and the degradation rate also given, just as in the case of the pre-mRNA, in units of 
the transcription rate relative to the splicing rate.

The analytical solutions to the dimensionless system reveal how the pre-mRNA 
and mRNA concentrations evolve over time towards their steady states. The 
exponential terms highlight the system's inherent timescales: $u^{\ast}$ relaxes 
to its steady state independently, while $s^{\ast}$’s dynamics are coupled to 
both $u^{\ast}$ and its own degradation rate. While this system is extremely
simplified relative to well-known biology, as a stylized model it still
indicates the interplay among transcription, splicing, and degradation rates.

While it is trivial to solve this system analytically, in more general cases this
 will not be possible. 
So we will proceed to simulate the system to characterize the possible solution 
sets and then perform probabilistic inference on the latent variables.

## Data

If we are just given samples of pre-mRNA and mRNA counts without additional
information about their sampling times, we can describe the data set 
$\mathcal{D}$ consisting of pre-mRNA and mRNA counts for a number 
of genes $G$ across a number of cells $N$

$$
\mathcal{D} = \left\{ (u_{ij}, s_{ij}) \mid i \in \{1, \ldots, G\}, j \in \{1, \ldots, N\} \right\}
$$

where

- $u_{ij}$ represents the count of pre-mRNA for gene $i$ in cell $j$,
- $s_{ij}$ represents the count of mRNA for the same gene $i$ in the same cell $j$,
- $G$ is the total number of genes in the study,
- $N$ is the total number of cells sampled.

Roughly speaking, the inference problem involves

1. **Normalization**: Transforming the observed counts $u_{ij}$ and $s_{ij}$ to 
   their dimensionless counterparts $u^{\ast}_{ij}$ and $s^{\ast}_{ij}$ using 
   any reasonable concentration scale $U_0$.
2. **Model Fitting**: Applying statistical methods to estimate the latent variables 
   $\Theta = \left( \gamma^{\ast}, u^{\ast}_0, s^{\ast}_0, t^{\ast} \right)$ that best 
   fit the observed data.
3. **Evaluation**: Assessing the fit of the model and the estimated parameters'
   biological plausibility and consistency across different genes and cells.

This representation of the dataset and the associated objective of 
statistical learning provides a solid foundation for applying probabilistic
modeling techniques to calibrate an inference procedure andevaluate the 
plausibility of this model in such a manner that we can eventually compare 
multiple candidate models to one another.

## Simulation

We will primarily focus on simulating the system, since this will generalize
to more complicated models. However, we will confirm our simulations for this
first and simplest model recapitulate its analytical solution derived above.

## Inference

Now that we have illustrated how to simulate the system, we can proceed to
define a probabilistic model and perform inference on its latent variables.
